Modelling the Trojan Asteroids

### 2020-04-24 Report/draft1.tex
periodicity of z perturbations

\section{Methodology}
This system of coupled first-order ordinary differential equations (ODEs) was solved using the scipy solve\_ivp function. The time span was taken as 100 orbits (with 100 points sampled per orbit) unless otherwise stated; this corresponds to 1185 Earth years. Rescaled solar system units are used for mathematical ease, so distances are measured in astronomical units (AU), time in earth years and mass in multiples of the solar mass, and to prevent floating point overflows due to the magnitude of the quantities considered in SI units.

@@

\subsection{Integration Method}
%Put this with theory as "numerical methods"?
%Then all theory could be in methodology?
Within solve\_ivp, the default solver is RK45 (an explicit Runge-Kutta method of order 5(4) \cite{Dormand1980}) and gives a deviation in asteroid position (from the Lagrange point) in the order of $ 10^{-4}$ AU in the rotating frame over 50 years. This is larger than expected, indicating the system of equations requires an unreasonably small step size for numerical stability with respect to this method, even in regions where the solution curve is smooth \cite{Lambert1991}. This suggests the system is stiff, and solvers designed for this typically do more work per step, allowing them to take much larger steps, and have improved numerical stability compared to the non-stiff solvers such as RK45 \cite{Byrne1987}.

@@

\subsection{Performance} \label{Performance}
Sampling 100 points per orbit for 100 orbits takes a mean time of $16.97 \pm 34$ \si{\milli\second}, with sub-linear scaling for sampling rate and orbit number up to the array memory limit, achieved through the optimised integration routines within solve\_ivp.

@@

In the stationary frame, this wander from the (now moving) Lagrange point is depicted in Figure \ref{fig:greeksdeviationstationaryframe}. The deviation oscillates with a magnitude of $9.10 \times 10^{-2}$ AU, and a period of $148 \pm 4$ years, where the error was determined by the width of the Fourier peak. This is modulated with a faster oscillation component of $11.85 \pm 0.27$ years, equivalent to the orbital period of the asteroids.

@@

\subsection{Perturbations in z-direction} \label{3D}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{Figures/wanderwithplanetmass_p5b}
	\caption{Variation of maximum wander with planetary mass. Note the well defined quadratic trend up to $ M_{p} = 0.026 M_{s}$, and instability beyond this.}
	\label{fig:planetmass}
\end{figure}

@@

\begin{table}[ht]
	\caption{Program files used to generate results in each section, with a description of their role.}
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		\textbf{Section} & \textbf{Description} & \textbf{Code File} \\
		\hline \hline
		\ref{Performance} & ODE solver performance & \textit{lagrange\_stability} \\
		\hline
		\ref{unperturbed} & Wander from unperturbed initial conditions & \textit{lagrange\_stability} \\
		\hline
		\ref{unperturbed} & Animations of unperturbed orbits & \textit{stat\_frame\_video} \\
		\hline
		\ref{perturbed}& Perturbations restricted to the xy plane & \textit{perturbation\_2D} \\
		\hline
		\ref{orbits}& Plotting of Orbit types in xy plane & \textit{perturbation\_2D} \\
		\hline
		\ref{3D}& Perturbations in the z-direction & \textit{perturbation\_3D} \\
		\hline
		\ref{planet}& Variation of planetary mass & \textit{planet\_mass\_variation} \\
		\hline
	\end{tabular}
	\label{Codefiles}
\end{table}

>>>>

\section{Methodology}
This system of coupled first-order ordinary differential equations (ODEs) was solved using the scipy \textit{solve\_ivp} function. The time span was taken as 100 orbits (with 100 points sampled per orbit) unless otherwise stated; this corresponds to 1185 Earth years. Rescaled solar system units are used for mathematical ease, so distances are measured in astronomical units (AU), time in earth years and mass in multiples of the solar mass, and to prevent floating point overflows due to the magnitude of the quantities considered in SI units.

@@

\subsection{Integration Method}
%Put this with theory as "numerical methods"?
%Then all theory could be in methodology?
Within \textit{solve\_ivp}, the default solver is RK45 (an explicit Runge-Kutta method of order 5(4) \cite{Dormand1980}) and gives a deviation in asteroid position (from the Lagrange point) in the order of $ 10^{-4}$ AU in the rotating frame over 50 years. This is larger than expected, indicating the system of equations requires an unreasonably small step size for numerical stability with respect to this method, even in regions where the solution curve is smooth \cite{Lambert1991}. This suggests the system is stiff, and solvers designed for this typically do more work per step, allowing them to take much larger steps, and have improved numerical stability compared to the non-stiff solvers such as RK45 \cite{Byrne1987}.

@@

\subsection{Performance} \label{Performance}
Sampling 100 points per orbit for 100 orbits takes a mean time of $16.97 \pm 34$ \si{\milli\second}, with sub-linear scaling for sampling rate and orbit number up to the array memory limit, achieved through the optimised integration routines within \textit{solve\_ivp}.

\subsection{Fourier Analysis} \label{Fourier}
For periodic oscillations/orbits, the Fast Fourier Transform (FFT) implemented in \textit{scipy} is used to obtain values for period, with peak finding in \textit{scipy.signal} identifying the exact values. The errors estimated using the associated peak widths of each frequency component.

@@

In the stationary frame, this wander from the (now moving) Lagrange point is depicted in Figure \ref{fig:greeksdeviationstationaryframe}. The deviation oscillates with a magnitude of $9.10 \times 10^{-2}$ AU, and a period of $148 \pm 4$ years. This is modulated with a faster oscillation component of $11.85 \pm 0.27$ years, equivalent to the orbital period of the asteroids.

@@

\subsection{Perturbations in Z-direction} \label{3D}
Perturbations in the Z-direction are aligned with angular velocity vector for rotating frame, so experience no virtual forces in this direction, and oscillate under the influence of gravity alone. Considering small perturbations in the z direction (so that the distance from the CoM can be considered constant), the period of such oscillations tends to the orbital period of the rotating frame, under Newtonian gravity.

These are measured to be $11.85 \pm 0.48$ years, in strong agreement with the theoretical predictions.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{Figures/wanderwithplanetmass_p5b}
	\caption{Variation of maximum wander with planetary mass, under a 0.01 AU radial perturbation and 0.01 AU/year tangential boost. Note the well defined quadratic trend up to $ M_{p} = 0.026 M_{s}$, and instability beyond this.}
	\label{fig:planetmass}
\end{figure}

@@

\begin{table}[ht]
	\caption{Program files used to generate results in each section, with a description of their role.}
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		\textbf{Section} & \textbf{Description} & \textbf{Code File} \\
		\hline \hline
		\ref{Performance} & ODE solver performance & \textit{lagrange\_stability} \\
		\hline
		\ref{Fourier} & Fourier Analysis & \textit{lagrange\_stability} \\
		\hline
		\ref{unperturbed} & Wander from unperturbed initial conditions & \textit{lagrange\_stability} \\
		\hline
		\ref{unperturbed} & Animations of unperturbed orbits & \textit{stat\_frame\_video} \\
		\hline
		\ref{perturbed}& Perturbations restricted to the xy plane & \textit{perturbation\_2D} \\
		\hline
		\ref{orbits}& Plotting of Orbit types in xy plane & \textit{perturbation\_2D} \\
		\hline
		\ref{3D}& Perturbations in the z-direction & \textit{perturbation\_3D} \\
		\hline
		\ref{planet}& Variation of planetary mass & \textit{planet\_mass\_variation} \\
		\hline
	\end{tabular}
	\label{Codefiles}
\end{table}
